Elaboration with First-Class Implicit Function Types

### 2020-02-05 paper.tex
tidy code examples

%% \usepackage{fontspec}
%% \fontspec[
%%   Path = fonts/,
%%   BoldFont = *-Bold,
%%   ItalicFont = *-Oblique,
%%   BoldItalicFont = *-BoldOblique]
%%   {DejaVuSansMono}
\setmonofont[Colour=blue,Scale=0.8]{DejaVuSansMono}
\usepackage{fancyvrb}
\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}

%% \newcommand{\blue}[1]{\textcolor{blue}{#1}}
%% \newcommand{\red}[1]{\textcolor{red}{#1}}
%% \newcommand{\green}[1]{\textcolor{ForestGreen}{#1}}
%% \newcommand{\purple}[1]{\textcolor{purple}{#1}}
%% \newcommand{\ASet}{\blue{Set}}
%% \newcommand{\CDefinition}{\blue{Definition}}
%% \newcommand{\CType}{\blue{Type}}
%% \newcommand{\Hforall}{\blue{forall}}
%% \newcommand{\Cforall}{\blue{forall}}
%% \newcommand{\HStar}{\blue{*}}

\newcommand{\slet}{\boldsymbol{\mathsf{let}}}
\renewcommand{\sin}{\boldsymbol{\mathsf{in}}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\To}{\Rightarrow}
\newcommand{\p}{\mathsf{p}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,#1\,#2\,#3}
\newcommand{\unify}{\mathsf{unify}}
\newcommand{\fail}{\mathsf{fail}}

\newcommand{\echeckblank}{\llbracket\blank\rrbracket\!\Downarrow}
\newcommand{\einferblank}{\llbracket\blank\rrbracket\!\Uparrow}
\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,#2\,#3\,#4}
\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,#2\,#3\,#4}
\newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
\newcommand{\einfert}[2]{\llbracket#1\rrbracket\!\Uparrow\,#2}
\newcommand{\edo}{\boldsymbol{\mathsf{do}}}
\newcommand{\ereturn}{\boldsymbol{\mathsf{return}}}
\newcommand{\eif}{\boldsymbol{\mathsf{if}}}
\newcommand{\ethen}{\boldsymbol{\mathsf{then}}}
\newcommand{\eelse}{\boldsymbol{\mathsf{else}}}
\newcommand{\ecase}{\boldsymbol{\mathsf{case}}}
\newcommand{\eof}{\boldsymbol{\mathsf{of}}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\einsert}[5]{\mathsf{insert}\,#1\,#2\,#3\,#4\,#5}

@@

\begin{abstract}
Implicit functions are dependently typed functions, such that arguments are
provided (by default) by inference machinery instead of programmers of the
surface language. Implicit functions in Agda are an archetypal example. In the
Haskell language as implemented by the Glasgow Haskell Compiler (GHC),
polymorphic types are another example. Implicit function types are
\emph{first-class} if they are treated as any other type in the surface
language. This is true in Agda and partially true in GHC. Inference and
elaboration in the presence of first-class implicit functions poses a challenge;
in the context of GHC and ML-like languages, this has been dubbed
``impredicative instantiation'' or ``impredicative inference''. We propose a new
framework for elaborating first-class implicit functions, which is applicable
for full dependent type theories and compares favorably to prior solutions in
terms of power, generality and conceptual simplicity. We build atop Norell's
bidirectional elaboration algorithm for Agda, and note that the key issue is
incomplete information about insertions of implicit abstractions and
applications. We make it possible to track and refine information related to
such insertions, by adding a new function type to a core Martin-L\"of type
theory, which supports strict (definitional) currying. This allows us to
represent undetermined domain arities of implicit function types, and we can
decide at any point during elaboration whether implicit abstractions should be
inserted.
\end{abstract}

@@

\textbf{Implicit function types} are a common mechanism for conveying to the
compiler that particular function arguments should be inferred by default. In
Agda and Coq, one can use bracketed function domains for this purpose:
\begin{Verbatim}[commandchars=\\\{\}]
          -- Agda                     (* Coq *)
          id : \{A : Set\} → A → A      Definition id \{A : Type\}(x : A) := x.
          id x = x
\end{Verbatim}
In GHC, one can use \texttt{forall} to define implicit function
types\footnote{This notation requires language extensions
  \texttt{KindSignatures} and \texttt{RankNTypes}; one could also write the type
  \texttt{a -> a} and GHC would silently insert the quantification.}
\begin{Verbatim}[commandchars=\\\{\}]
                         id :: forall (a :: *). a -> a
                         id x = x
\end{Verbatim}

In all of the above cases, if we apply \texttt{id} to an argument, the implicit
type argument is provided by elaboration. For example, in Agda, \texttt{id true}
is elaborated to \texttt{id \{Bool\} true}, and analogously in GHC and Coq. In all
three systems, there is also a way to explicitly specify implicit arguments: in
Agda we may put arguments in brackets as we have seen, in Coq we can prefix a
name with \texttt{@} to make every implicit argument explicit, as in \texttt{@id
  bool true}, and in GHC we can enable the language extension
\texttt{TypeApplications} and write \texttt{id @Bool True}.

Implicit functions are \textbf{first-class} if they can be manipulated like any
other type. Coq is an example for a system where this is \emph{not} the case.
In Coq, the core language does not have an actual implicit function type,
instead, implicitness is tied to particular \emph{names}, and while we can write
\texttt{List (forall \{A : Type\}, A -> A)}, the brackets here are simply
ignored by Coq and we get a plain function type. For example, Coq accepts the
following definition:
\begin{Verbatim}[commandchars=\\\{\}]
      Definition poly : forall (f : forall \{A : Type\}, A -> A), bool * nat :=
        fun f => (f bool true, f nat 0).
\end{Verbatim}
This is a higher-rank polymorphic function which returns a pair. Note that
\texttt{f} is applied to two arguments, because the implicitness in
\texttt{forall \{A : Type\}, A -> A} is silently dropped.

In GHC Haskell, \texttt{forall} types are more flexible. We can write
the following, with \texttt{RankNTypes} enabled:
\begin{Verbatim}[commandchars=\\\{\}]
                     poly :: (forall a. a -> a) -> (Bool, Int)
                     poly f = (f True, f 0)
\end{Verbatim}
However, polymorphic types are only supported in function domains and as fields
of algebraic data constructors. We cannot instantiate an arbitrary type
parameter to a \texttt{forall}, as in \texttt{[forall a. a -> a]} for a list
type with polymorphic elements. While this type is technically allowed by the
\texttt{ImpredicativeTypes} language extension, as of GHC 8.8 this extension is
deprecated and is not particularly usable in practice.

In Agda, implicit functions are truly a first-class notion, and we may have
\texttt{List (\{A : Set\} → A → A)} without issue. However, Agda's elaboration
still has limitations when it comes to handling implicit functions. Assume that
we have \texttt{[]} for the empty list and \texttt{\_∷\_} for list extension,
and consider the following code:
\begin{Verbatim}[commandchars=\\\{\}]
                     polyList : List (\{A : Set\} → A → A)
                     polyList = (λ x → x) ∷ []
\end{Verbatim}
Agda 2.6.0.1 does not accept this. However, it does accept \texttt{polyList = (λ
  \{A\} x → x) ∷ []}. The issue is the following. Agda first \emph{infers} a
type for \texttt{(λ x → x) ∷ []}, then tries to unify the inferred type with the
given \texttt{List (\{A : Set\} → A → A)} annotation. However, when Agda
elaborates \texttt{λ x → x}, it does not yet know anything about the element
type of the list; it is an undetermined unification variable. Hence, Agda does
not know whether it should insert an extra \texttt{λ\{A\}} or not. If the
element type is later found to be an implicit function, then it should, otherwise it
should not. To solve this conundrum, Agda simply assumes that any unknown type
is \emph{not} an implicit function type, and elects to not insert a lambda. This
assumption is often correct, but sometimes --- as in the current case --- it is not.

@@

The solution presented in this paper is to gradually accummulate information
about implicit insertions, and to have a setup where insertions can be refined
and performed at any time after a particular expression is elaborated. In the
current example, our algorithm wraps \texttt{λ x → x} in an implicit lambda with
unknown arity, and the domain type is later refined to be \texttt{A : Set} when
the inferred type is unified with the annotation.

@@

\subsubsection{Note on Terminology}

We prefer to avoid the term ``impredicative inference'' in order to avoid
confusion with impredicativity in type theory. The two notions historically
coincided, but has since diverged to the point of being orthogonal. In type
theory, impredicativity is a property of a universe, i.e.\ closure of a universe
under arbitrary products. In the type inference literature, impredicativity
means the ability to instantiate type variables and metavariables to polymorphic
types. In particular, we have that
\begin{itemize}
  \item Agda has type-theory-predicative universes, but implements
    type-inference-impredicative elaboration with first-class implicit function
    types.
  \item Coq has type-theory-impredicative \texttt{Prop} universe (and
    optionally also \texttt{Set}), but implements
    type-inference-predicative elaboration, because of the lack of implicit
    function types.
  \item GHC is type-theory-impredicative with \texttt{RankNTypes} enabled
    and \texttt{ImpredicativeTypes} disabled, as we have
    \texttt{(forall (a :: *). a -> a) :: *}.
\end{itemize}

@@

\begin{example}
The core syntax is quite expressive as a programming language, thanks to
let-definitions and the type-in-type rule which allows Church-encodings of a
large class of inductive types. For example, the following term computes a list
of types, by mapping over the list $cons\,\U\,(cons\,\U\,nil)$.
\begin{alignat*}{3}
  & \slet\,List : \U\to\U\\
  & \hspace{1em}= \lambda\,A.\,(L : \U)\to(A\to L\to L)\to L\to L\,\,\sin\\
  & \slet\,map : \{A\,B : \U\}\to (A \to B) \to List\,A \to List\,B\\
  & \hspace{1em}=
  \lambda\,\{A\}\,\{B\}\,f\,as\,L\,cons\,nil.\,as\,L\,(\lambda\,a.\,cons\,(f\,a))\,nil\,\,\sin\\
  & map\,\{\U\}\,\{\U\}\,(\lambda\,A.\, A \to A)\,(\lambda\,L\,cons\,nil.\,cons\,\U\,(cons\,\U\,nil))
\end{alignat*}
In the core syntax, all applications must be explicit, so using implicit
function type for $map$ above does not make much difference in optics.
\end{example}

@@

We define a helper function which inserts implicit applications around a core
term. For example, if we have a defined name $id$ with type $\{A : \U\}\to A \to
A$ occurring in the surface program, we usually want to expand $id$ to
$id\,\{\alpha\}$, where $\alpha$ is a fresh metavariable.
\begin{mathpar}
\inferrule*{ins \in \{\true,\,\false\} \\ \Theta|\Gamma\vdash t : A}
           {\einsert{ins}{\Theta}{\Gamma}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
             (\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t' : A')\}}
\end{mathpar}
We have an additional $ins \in \{\true,\,\false\}$ parameter, which simply
toggles whether any insertion is to be performed. We will use this in the
definition of elaboration. Insertion is defined by recursion on the $A$ core
type, as follows.
\begin{alignat*}{3}
  &\einsert{\false}{\Theta}{\Gamma}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
  &\einsert{\true\,}{\Theta}{\Gamma}{t}{(\{x : A\}\to B)} && :\equiv&&
       \slet\,(\Theta',\,\theta,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
  & && &&\sin\,\einsert{\true}{\Theta'}{(\Gamma[\theta])}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
  &\einsert{\true\,}{\Theta}{\Gamma}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
\end{alignat*}

@@

We assume that there is a partial unification procedure, which returns a unifying
metasubstitution on success. We only have \emph{homogeneous} unification, i.e.\ the
two terms to be unified must have the same type. The specification is as follows:
\begin{mathpar}
\inferrule*{\Theta|\Gamma\vdash t : A \\ \Theta|\Gamma\vdash u : A}
           {\unify\,t\,u \in \{(\Theta',\,\theta)\,|\,(\theta : \Theta' \To \Theta)
             \,\land\,(\Theta'|\Gamma[\theta]\vdash
             t[\theta] \equiv u[\theta] : A[\theta])\}\,\cup\,\{\fail\}}
\end{mathpar}

@@

\todo{example for meta solution, more references}

@@

\subsubsection{On soundness, completeness, and other properties}
We briefly discuss what we can and cannot say about the currently presented
elaborator. First, elaboration is \emph{sound} in the sense that it never
produces malformed core syntax.

@@

We can still say something about what kind of work our elaborator does. For
this, we consider a translation from core terms to surface terms, the obvious
``forgetful'' translation, which maps terms to surface counterparts. Now, this
is an ``evil'' construction on core terms, since it does not preserve
definitional equality. We shall only use this evil notion in the following
statement.

@@

>>>>

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{cleveref}

\newcommand{\slet}{\boldsymbol{\mathsf{let}}}
\renewcommand{\sin}{\boldsymbol{\mathsf{in}}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\To}{\Rightarrow}
\newcommand{\p}{\mathsf{p}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,#1\,#2\,#3}
\newcommand{\unify}{\mathsf{unify}}
\newcommand{\fail}{\mathsf{fail}}
\newcommand{\kw}[1]{{\mathsf{#1}}}

\newcommand{\echeckblank}{\llbracket\blank\rrbracket\!\Downarrow}
\newcommand{\einferblank}{\llbracket\blank\rrbracket\!\Uparrow}
\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,#2\,#3\,#4}
\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,#2\,#3\,#4}
\newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
\newcommand{\einfert}[2]{\llbracket#1\rrbracket\!\Uparrow\,#2}
\newcommand{\edo}{\boldsymbol{\mathsf{do}}}
\newcommand{\ereturn}{\boldsymbol{\mathsf{return}}}
\newcommand{\eif}{\boldsymbol{\mathsf{if}}}
\newcommand{\ethen}{\boldsymbol{\mathsf{then}}}
\newcommand{\eelse}{\boldsymbol{\mathsf{else}}}
\newcommand{\ecase}{\boldsymbol{\mathsf{case}}}
\newcommand{\eof}{\boldsymbol{\mathsf{of}}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\einsert}[5]{\mathsf{insert}\,#1\,#2\,#3\,#4\,#5}
\newcommand{\Set}{\mathsf{Set}}

@@

\begin{abstract}
Implicit functions are dependently typed functions, such that arguments are
provided (by default) by inference machinery instead of programmers of the
surface language. Implicit functions in Agda are an archetypal example. In the
Haskell language as implemented by the Glasgow Haskell Compiler (GHC),
polymorphic types are another example. Implicit function types are
\emph{first-class} if they are treated as any other type in the surface
language. This holds in Agda and partially holds in GHC. Inference and
elaboration in the presence of first-class implicit functions poses a challenge;
in the context of GHC and ML-like languages, this has been dubbed
``impredicative instantiation'' or ``impredicative inference''. We propose a new
framework for elaborating first-class implicit functions, which is applicable
for full dependent type theories and compares favorably to prior solutions in
terms of power, generality and conceptual simplicity. We build atop Norell's
bidirectional elaboration algorithm for Agda, and note that the key issue is
incomplete information about insertions of implicit abstractions and
applications. We make it possible to track and refine information related to
such insertions, by adding a new function type to a core Martin-L\"of type
theory, which supports strict (definitional) currying. This allows us to
represent undetermined domain arities of implicit function types, and we can
decide at any point during elaboration whether implicit abstractions should be
inserted.
\end{abstract}

@@

\textbf{Implicit function types} are a common mechanism for conveying to the
compiler that particular function arguments should be inferred by default. In
Agda and Coq, one can use bracketed function domains for this purpose:
\begin{alignat*}{3}
  & id : \{A : \kw{Set}\}\to A \to A \hspace{5em} \kw{Definition}\,id\,\{A : \kw{Type}\}(x : A) := x.\\
  & id\,x = x
\end{alignat*}
In GHC, one can use $\kw{forall}$ to define implicit function
types\footnote{This notation requires language extensions
  $\mathsf{KindSignatures}$ and $\mathsf{RankNTypes}$; one could also write the type
  $a \to a$ and GHC would silently insert the quantification.}
\begin{alignat*}{3}
  & id :: \kw{forall}\,(a :: \kw{*}).\,a \to a\\
  & id\,x = x
\end{alignat*}
In all of the above cases, if we apply $id$ to an argument, the implicit type
argument is provided by elaboration. For example, in Agda, $id\,true$ is
elaborated to $id\,Bool\,true$, and analogously in GHC and Coq. In all three
systems, there is also a way to explicitly specify implicit arguments: in Agda
we may put arguments in brackets as we have seen, in Coq we can prefix a name
with $\kw{@}$ to make every implicit argument explicit, as in
$\kw{@}id\,bool\,true$, and in GHC we can enable the language extension
$\mathsf{TypeApplications}$ and write $id\,\kw{@}Bool\,True$.

Implicit functions are \textbf{first-class} if they can be manipulated like any
other type. Coq is an example for a system where this is \emph{not} the case.
In Coq, the core language does not have an actual implicit function type,
instead, implicitness is tied to particular \emph{names}, and while we can write
$list\,(\kw{forall}\,\{A : \kw{Type}\},\,A\to A)$ for a list type of polymorphic
functions, the brackets here are simply ignored by Coq and we get a plain
function type. For example, Coq accepts the following definition:
\begin{alignat*}{3}
  & \kw{Definition}\,poly\,: \kw{forall}\,(f : \kw{forall}\,\{A : \kw{Type}\},\,A\to A),\,bool\,*\,nat :=\\
  & \quad \kw{fun}\,f \Rightarrow (f\,bool\,true,\,f\,nat\,0)
\end{alignat*}
This is a higher-rank polymorphic function which returns a pair. Note that $f$
is applied to two arguments, because the implicitness in $\kw{forall}\,\{A :
\kw{Type}\},\,A\to A$ is silently dropped.

In GHC Haskell, $\kw{forall}$ types are more flexible. We can write the
following, with $\kw{RankNTypes}$ enabled:
\begin{alignat*}{3}
  & poly :: (\kw{forall}\,a.\,a \to a) \to (Bool,\,Int) \\
  & poly\,f = (f\,True,\,f\,0)
\end{alignat*}
However, polymorphic types are only supported in function domains and as fields
of algebraic data constructors. We cannot instantiate an arbitrary type
parameter to a $\kw{forall}$, as in $[\kw{forall}\,a.\,a\to a]$ for a list type
with polymorphic elements. While this type is technically allowed by the
$\kw{ImpredicativeTypes}$ language extension, as of GHC 8.8 this extension is
deprecated and is not particularly usable in practice.

In Agda, implicit functions are truly a first-class notion, and we may have
$List\,(\{A : \kw{Set}\}\to A \to A)$ without issue. However, Agda's elaboration
still has limitations when it comes to handling implicit functions. Assume that
we have $[]$ for the empty list and $\blank::\blank$ for list extension,
and consider the following code:
\begin{alignat*}{3}
  & polyList : List\,(\{A : \kw{Set}\}\to A \to A)\\
  & polyList = (\lambda\,x \to x)\,::\,[]
\end{alignat*}
Agda 2.6.0.1 does not accept this. However, it does accept $polyList =
(\lambda\,\{A\}\,x \to x)\,::\,[]$. The issue is the following. Agda first
\emph{infers} a type for $(\lambda\,x \to x)\,::\,[]$, then tries to unify the
inferred type with the given $List\,(\{A : \kw{Set}\}\to A \to A)$
annotation. However, when Agda elaborates $\lambda\,x \to x$, it does not yet
know anything about the element type of the list; it is an undetermined
unification variable. Hence, Agda does not know whether it should insert an
extra $\lambda\,\{A\}$ or not. If the element type is later found to be an
implicit function, then it should, otherwise it should not. To solve this
conundrum, Agda simply assumes that any unknown type is \emph{not} an implicit
function type, and elects to not insert a lambda. This assumption is often
correct, but sometimes --- as in the current case --- it is not.

@@

The solution presented in this paper is to gradually accummulate information
about implicit insertions, and to have a setup where insertions can be refined
and performed at any time after a particular expression is elaborated. In the
current example, our algorithm wraps $\lambda\,x\to x$ in an implicit lambda
with unknown arity, whose domain is later refined to be $A : \kw{Set}$ when the
inferred type is unified with the annotation.

@@

\subsubsection{Note on terminology}

We prefer to avoid the term ``impredicative inference'' in order to avoid
confusion with impredicativity in type theory. The two notions historically
coincided, but has since diverged to the point of being orthogonal. In type
theory, impredicativity is a property of a universe, i.e.\ closure of a universe
under arbitrary products. In the type inference literature, impredicativity
means the ability to instantiate type variables and metavariables to polymorphic
types. In particular, we have that
\begin{itemize}
  \item Agda has type-theory-predicative universes, but implements
    type-inference-impredicative elaboration with first-class implicit function
    types.
  \item Coq has type-theory-impredicative $\kw{Prop}$ universe (and optionally
    also $\kw{Set}$), but implements type-inference-predicative elaboration,
    because of the lack of implicit function types.
  \item GHC is type-theory-impredicative with $\kw{RankNTypes}$ enabled and
    $\kw{ImpredicativeTypes}$ disabled, as we have $(\kw{forall}\,(a :: *).\, a
    \to a) :: *$.
\end{itemize}

@@

\begin{example}
The core syntax is quite expressive as a programming language, thanks to
let-definitions and the type-in-type rule which allows Church-encodings of a
large class of inductive types. For example, the following term computes a list
of types, by mapping over the list $cons\,\U\,(cons\,\U\,nil)$.
\begin{alignat*}{3}
  & \slet\,List : \U\to\U\\
  & \hspace{1em}= \lambda\,A.\,(L : \U)\to(A\to L\to L)\to L\to L\,\,\sin\\
  & \slet\,map : \{A\,B : \U\}\to (A \to B) \to List\,A \to List\,B\\
  & \hspace{1em}=
  \lambda\,\{A\}\,\{B\}\,f\,as\,L\,cons\,nil.\,as\,L\,(\lambda\,a.\,cons\,(f\,a))\,nil\,\,\sin\\
  & map\,\{\U\}\,\{\U\}\,(\lambda\,A.\, A \to A)\,(\lambda\,L\,cons\,nil.\,cons\,\U\,(cons\,\U\,nil))
\end{alignat*}
In the core syntax, all applications must be explicit, so using implicit
function type for $map$ above does not make much difference in notation.
\end{example}

@@

We define a helper function which inserts implicit applications around a core
term. For example, if we have a defined name $id$ with type $\{A : \U\}\to A \to
A$ occurring in the surface program, we usually want to expand $id$ to
$id\,\{\alpha\}$, where $\alpha$ is a fresh metavariable.
\begin{mathpar}
\inferrule*{ins \in \{\true,\,\false\} \\ \Theta|\Gamma\vdash t : A}
           {\einsert{ins}{\Theta}{\Gamma}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
             (\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t' : A')\}}
\end{mathpar}
We have an additional $ins \in \{\true,\,\false\}$ parameter, which simply
toggles whether any insertion is to be performed. We will use this in the
definition of elaboration. Insertion is defined by recursion on $ins$ and the
$A$ type, as follows.
\begin{alignat*}{3}
  &\einsert{\false}{\Theta}{\Gamma}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
  &\einsert{\true\,}{\Theta}{\Gamma}{t}{(\{x : A\}\to B)} && :\equiv&&
       \slet\,(\Theta',\,\theta,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
  & && &&\sin\,\einsert{\true}{\Theta'}{(\Gamma[\theta])}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
  &\einsert{\true\,}{\Theta}{\Gamma}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
\end{alignat*}

@@

We assume that there is a unification procedure, which returns a unifying
metasubstitution on success. We only have \emph{homogeneous} unification,
i.e.\ the two terms to be unified must have the same type. The specification is
as follows:
\begin{mathpar}
\inferrule*{\Theta|\Gamma\vdash t : A \\ \Theta|\Gamma\vdash u : A}
           {\unify\,t\,u \in \{(\Theta',\,\theta)\,|\,(\theta : \Theta' \To \Theta)
             \,\land\,(\Theta'|\Gamma[\theta]\vdash
             t[\theta] \equiv u[\theta] : A[\theta])\}\,\cup\,\{\fail\}}
\end{mathpar}

@@

\todo{example for meta solution, more references, describe at least pattern unif and pruning briefly}

@@

\subsubsection{On soundness, completeness, and other properties}
We briefly discuss what we can and cannot say about the presented
elaborator. First, elaboration is \emph{sound} in the sense that it never
produces malformed core syntax.

@@

We can still say something about the behavior of our elaborator. For this, we
consider a translation from core terms to surface terms, the obvious forgetful
translation, which maps core terms to surface counterparts. Now, this is an
``evil'' construction on core terms, since it does not preserve definitional
equality. We shall only use this evil notion in the following statement.

@@

\todo{Say something about termination of elab?}
